---
title: "geneBasis, mouseEmbryo, extended workflow."
author:
  - Alsu Missarova
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
    number_sections: yes
    toc: yes  
  BiocStyle::pdf_document: default
package: geneBasisR
vignette: >
  %\VignetteIndexEntry{geneBasis_mouseEmbryo_extended}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

library(geneBasisR)
library(SingleCellExperiment)
library(tibble)
library(ggplot2)
library(wesanderson)
library(ggcorrplot)
library(ggpubr)

```

# Introduction

`geneBasis` is an iterative greedy approach for gene selection. It strives a gene selection that will preserve transcriptional relationships between cells, represented as kNN-graphs.


Here we exemplify how `geneBasis` can be utilized for library design. First, we design a library of designated size from scratch. We then estimate the quality of designed library and provide a workflow that can be used to investigate cells and/or genes that are not explained with the selected library. Finally, we provide the estimation of the redundancy of the library for cell type mapping.

# Load data

We will be working with mouse embryo, E8.5. Initial filtering for uninteresting genes was already performed (to see how genes can be pre-filtered, see Vignette for spleen).

```{r load-data}

data("sce_mouseEmbryo", package = "geneBasisR")

# Fetch meta-data and umap coordinates, merge together
meta = as.data.frame(colData(sce_mouseEmbryo))
umaps = as.data.frame(reducedDim(sce_mouseEmbryo, "UMAP"))
umaps = rownames_to_column(umaps, var = "cell")
meta = merge(meta, umaps)

# Fetch colour scheme for celltypes from meta-data
celltype_colors_df = unique(meta[, c("celltype" , "colour")])
celltype_colors = c(paste0("#",as.character(celltype_colors_df$colour)))
names(celltype_colors) = as.character(celltype_colors_df$celltype)

```

# Cell type composition for mouse embryo

Let's see which cell types (and in which quantities) we observe in mouse embryo.

```{r umap-ct, fig.wide = TRUE, fig.cap = "Cell type composition."}

# celltype composition
tab = as.data.frame(table(sce_mouseEmbryo$celltype))
colnames(tab) = c("celltype" , "n")
tab = tab[order(tab$n) , ]
tab$celltype = factor(tab$celltype , levels = tab$celltype)
p1 = ggplot(tab , aes(x = celltype , y = log2(n) , fill = celltype)) +
  geom_bar(stat = "identity" , position = "dodge") +
  scale_fill_manual(values = celltype_colors) +
  theme_classic() +
  theme(axis.text.x = element_blank()) +
  labs(y = "log2(# cells)", x = "Cell type")

# UMAP
p2 = ggplot(meta , aes(x = x , y = y , col = celltype)) +
  geom_point(size=1,alpha = .9) +
  scale_color_manual(values = celltype_colors) +
  theme_classic() +
  labs(x = "UMAP-1" , y = "UMAP-2") +
  ggtitle("UMAP")

# combine
p = ggarrange(p1,p2, common.legend = T)
p


```

# Select library of size 50

To select library, use `gene_search`. 

We will not print intermediate outputs for this run (by setting `verbose=FALSE`).
We also use `sample` as batch identification. 

```{r gene-search}
 
n_genes_total = 50
genes_stat = gene_search(sce_mouseEmbryo , n_genes_total = n_genes_total, batch = "sample", verbose = F)
genes = genes_stat$gene


```

# Evaluation of the library

We estimate the quality of the library using `evaluate_library`.

## Gene and cell scores convergence - as functions of n-genes

Here, to be less wordy, we refer to cell neighborhood preservation score as cell score; gene prediction score as gene score.

```{r cell-gene-score-stat, fig.wide = TRUE, fig.cap="Cell and gene scores - as functions of n-genes."}


stat = evaluate_library(sce_mouseEmbryo, genes, genes.all = rownames(sce_mouseEmbryo), batch = "sample", library.size_type = "series", n_genes.step = 10,
                        return.cell_score_stat = T, return.gene_score_stat = T, return.celltype_stat = T, verbose = F)

p1 = ggplot(stat$cell_score_stat , aes(x = n_genes , y = cell_score, fill = n_genes)) +
  geom_boxplot() +
  scale_fill_manual(values = wes_palette("Zissou1", length(levels(stat$cell_score_stat$n_genes)), type = "continuous")) +
  ylim(c(-.2,1)) +
  theme_classic() +
  labs(y = "Cell neighborhood preservation score" , x = "# genes") +
  theme(legend.position = "none")
p2 = ggplot(stat$gene_score_stat , aes(x = n_genes , y = gene_score, fill = n_genes)) +
  geom_boxplot() +
  scale_fill_manual(values = wes_palette("Zissou1", length(levels(stat$gene_score_stat$n_genes)), type = "continuous")) +
  ylim(c(0,1)) +
  theme_classic() +
  labs(y = "Gene prediction score" , x = "# genes") +
  theme(legend.position = "none")

p = ggarrange(p1,p2,ncol=2)
p



```

We suggest that these plots can be useful to make a decision at which number of genes cell and gene score distributions converge (more relevant for higher number of genes).

In our example, we observe that distributions do not converge yet and therefore we suggest that 50 genes is not above optimal size for the library.

## Cell type mapping 

Let's see how well we can map cell types using selected library.

### Confusion matrix

Cell type mapping for a single library can be computed separately using `get_celltype_mapping`. We also included plotting fucntion that returns heatmap for the confusion matrix: `plot_mapping_heatmap`.


```{r celltype-mapping-confusion, fig.wide = TRUE,  fig.height=10, fig.cap="Cell type confusion matrix."}


celltype_mapping = get_celltype_mapping(sce_mouseEmbryo , genes, batch = "sample", return.stat = F)
p = plot_mapping_heatmap(celltype_mapping$mapping, title = "Cell type confusion matrix")
p


```

### Cell type mapping accuracy - as a function of n-genes

```{r celltype-mapping-trend, fig.wide = TRUE, fig.cap="Cell type mapping accuracy - as a function of n-genes."}


pal = wes_palette("Zissou1" , length(levels(stat$celltype_stat)) , type = "continuous")
p = ggplot(stat$celltype_stat, aes( x = as.character( n_genes ), y = frac_correctly_mapped , col = "black")) +
  geom_point(size=1.5) +
  facet_wrap(~celltype) +
  #scale_color_manual(values = celltype_colors) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  theme_classic() +
  theme(legend.position = "none") +
  labs(y = "Fraction of cells mapped correctly", x = "# genes")
  ggtitle("Cell type mapping accuracy")
p

```

## Cell score - distribution per cell type 

Let's assess whether certain cell types are consistently 'under-preserved' by looking at the distribution of cell scores across cell types.

```{r cell-score-by-ct-mapping, fig.wide = TRUE,  fig.cap="Cell score per cell type."}


cell_score_stat = stat$cell_score_stat[stat$cell_score_stat$n_genes == n_genes_total , ] 
cell_score_stat = merge(cell_score_stat, meta) 

p = ggplot(cell_score_stat , aes(x = celltype , y = cell_score, fill = celltype)) + 
  geom_boxplot() + 
  scale_fill_manual(values = celltype_colors) + 
  theme_classic() + 
  labs(y = "Cell neighborhood preservation score" , x = "# genes") + 
  theme(legend.position = "none") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 
p 


``` 

We see that PGCs in average have lower scores. This is expected since we have only 8 PGC cells and our toy library is of very limited size. We also observe that a decent fraction of PGCs do not map correctly to the same cell type (Fig. 3). We therefore suggest to increase number of genes to be selected or manually add PGC-specific markers into the library (especially if detecting PGCs is of high interest).


### See which genes we do not predict well 

Now let's assess which genes can not be predicted well from k-NN graph constructed with selected library. 

We assign threshold for prediction score as 0.75 and look at the genes that have gene score < 0.75.

#### Co-expression 

Let's see if we observe a highly co-expressed gene set. The plotting function is included in the package: `plot_coexpression`.


```{r coexpression-lowly-predicted-genes, fig.wide = TRUE,  fig.cap="Co-expression within lowly predicted genes."} 

gene_score.thresh = 0.75 
gene_score_stat = stat$gene_score_stat[stat$gene_score_stat$n_genes == n_genes_total , ] 

genes_lowly_explained = gene_score_stat$gene[gene_score_stat$gene_score < gene_score.thresh] 

p = plot_coexpression(sce_mouseEmbryo , genes_lowly_explained ) 
p 


``` 

#### Average expression across cell types

A function to plot a heatmap of average expression across is built in the library: `plot_expression_heatmap`.

```{r avg-expression-heatmap-lowly-predicted-genes, fig.wide = TRUE, fig.cap="Average expression per cell type, lowly explained genes."}

p = plot_expression_heatmap(sce_mouseEmbryo, genes_lowly_explained, value.type = "mean") 
p 

```

We see that most of selected genes are cell state genes, expressed across most of the cell types. 

Interestingly, we also see that strong PGC marker Dppa3 is not explained well. This is consistent with the observation from the above. 
Additionally, blood progenitor marker F2rl2 is also not explained well, and this is also consistent with cell type confusion matrix from Fig. 3.

We suggest that this workflow can facilitate the understanding which genes are 'missing' from the library and whether it is desirable to include them. 

# Characterisation of library 

Let's actually look into what genes we select.

## What are the genes we are selecting - visual inspection

Let's plot UMAPs, colored by expression of the selected genes, to have an intuitive grasp of what kind of genes we select. UMAPs can be easily plotted with provided function `plot_umaps_w_counts`. 

```{r umaps-selected-genes, fig.wide = TRUE, fig.height=15, fig.cap="UMAPs of selected genes."}

p = plot_umaps_w_counts(sce_mouseEmbryo, genes)
p

```

Overall we see that most of the genes are expressed in some transcriptional regions but not others (i.e. like cell type markers behavior). 

Additionally, some genes seem to be expressed across the whole manifold and are representative of a cell state (Ube2c, Cdc20, Xist, etc).

## Average expression across cell types and co-expression 

Let's look a bit more systematically into how cell type specific are expressions of the selected genes.

```{r general-stat-heatmap-avg-expression, fig.wide = TRUE, fig.cap="Average expression per cell type."}

p = plot_expression_heatmap(sce_mouseEmbryo, genes, value.type = "mean") 
p 

``` 

## Estimation of redundancy 

Finally, let's estimate the redundancy of the library on gene/celltype level: for each gene, we temporarily remove it from the library and compare  the accuracies of cell type mappings with and without removed gene (per cell type). 

The function to get this stat is `get_redundancy_stat`. We also provide the function that will plot the results as a heatmap: `plot_redundancy_stat`. Red corresponds to cases where cell type mapping accuracy drops with the removal of the gene, green - where cell type mapping accuracy increases with the removal of the gene (that can occur for cell type that were not well mapped to begin with).

```{r redundancy-plot, fig.wide = TRUE, fig.cap="Redundancy heatmap."} 

redundancy_stat = get_redundancy_stat(sce_mouseEmbryo, genes, genes_to_assess = genes, batch = "sample") 
p = plot_redundancy_stat(redundancy_stat) 
p 

``` 

From this analysis, we can conclude the particular relevance of some genes for some cell types e.g. Mest for Blood progenitors 1, Hbb-y for Erythroid2.

This line of analysis also allows to estimate overall redundancy of the library which is useful when planning seq-FISH experiments.


#Session Info
```{r sessinf}
sessionInfo()
```
